---
title: "Heatmaps"
author: "Jessica Phung"
date: "February 21, 2016"
output: html_document
---

## The two steps are to combine two datasets and make a heatmap. The code below will combine two of the expression sets. If batch effects are present there should be 'clusters' of different shades of color. If there is a lack of batch effects most of the data will have a similar coloration and should look like a near unified shade.

## The code below will establish what each eset looks like. 
```{r}
library(Biobase)
heatmap(exprs(var.feset1[1:40]))
```

```{r, eval=FALSE}
heatmap(exprs(var.feset2[1:96]))

heatmap(exprs(var.feset3[1:50]))
```

##Combine two expression sets (var.functioneset1, var.feset2) then make a heatmap.
var.feset1.2<-combine(var.feset1,var.feset2)

heatmapvar.feset1.2<-heatmap(exprs(var.feset1.2[1:136]))

##Repeat steps for the other esets paired with eset1. 

var.feset1.3<-combine(var.feset1,var.feset3)

heatmapvar.feset1.3<-heatmap(exprs(var.feset1.3[1:90]))

## Pair up esets 2 and 3.

var.feset2.3<-combine(var.feset2,var.feset3)

heatmapvar.feset2.3<-heatmap(exprs(var.feset2.3[1:146]))

## The heatmaps should show up on the left under 'plots' and can be coverted to an image or pda. The batch effects should be very clear in the heatmaps. 







